Semantic Linearity in a Lazy Optimising Compiler

### 2025-07-10 icfp.tex
Clarify case expression WHNF judgment

%Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
A case expression drives evaluation by evaluating its scrutinee to Weak Head
Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
alternative whose pattern matches the result of evaluating the
scrutinee is taken\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}.
An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $(\lambda x.~e)$ 
or a datatype constructor application $(K~\ov{e})$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF. % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed and such resources will only be consumed when the sub-expressions themselves are evaluated.

As will be made clear in this section, we need to devise a specialized typing
for scrutinees that distinguishes between terms
in WHNF and terms that are not in WHNF.
Following the discussion on expressions in WHNF, we present a
typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$, and a rule for each
of the forms given above:
\[
    \TypeWHNFCons
\qquad
    \TypeWHNFLam
\]
This judgement differs from the main typing judgement in that (1) it only
applies to expressions in weak head normal form, and (2) it ``outputs'' (to the right of $\gtrdot$) a
disjoint set of linear environments ($\ov{\D_i}$), where each environment corresponds to the
linear resources used by a sub-expression of the WHNF expression.
%
To type a constructor application $K~\ov{e_\omega e_i}$, where $e_\omega$
are the unrestricted arguments and $e_i$ the linear arguments of the
constructor, we split the resources $\D$ into a disjoint set of resources
$\ov{\D_i}$ required to type each linear argument individually and return exactly
that split of the resources; the unrestricted $e_\omega$ expressions must be
typed on an empty linear environment. A lambda expression is typed with
the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
as there is always only a single sub-expression in lambda abstractions.

@@

We highlight that, when evaluating a case expression, computation only
happens when a scrutinee that is not in WHNF is evaluated to
WHNF. Otherwise, evaluation continues in the alternative by
substituting in the appropriate scrutinee expressions, but without
having performed any additional computation.
% (the scrutinee was already in weak head normal form).
%In short, no computation happens if the scrutinee is already in WHNF.
%
In terms of linearity, resources are consumed only if evaluation happens.
Therefore, resources used to type a scrutinee not in
WHNF will be consumed when the case is evaluated and will no longer be directly available in the case
alternatives. Conversely, when the scrutinee is already in WHNF, linear
resources required to type the scrutinee are still directly available in the
alternatives.
%
%%%%???
%%%%The linear resources used to type an expression in WHNF are exactly those which
%%%%occur to the right of $\gtrdot$ in the WHNF judgement shown above
%%%%(corresponding to the resources required to typecheck the lambda body or the
%%%%constructor arguments).

>>>>

%Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
A case expression drives evaluation by evaluating its scrutinee to Weak Head
Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
alternative whose pattern matches the result of evaluating the
scrutinee is taken\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}.
An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $(\lambda x.~e)$ 
or a datatype constructor application $(K~\ov{e})$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
(and they may capture linear resources).

A key detail is that when the scrutinee is already in WHNF, evaluation continues
in the alternative simply by substituting the pattern variables by the
scrutinee sub-expressions. No resource from the scrutinee can be consumed since
no computation happens at all.
% (the scrutinee was already in weak head normal form).
%In short, no computation happens if the scrutinee is already in WHNF.
%
% In light of this, our type system will need to distinguish between terms in WHNF and terms
% that are not in WHNF.
%
% TODO While it may seem ad-hoc to type an expression differently
%
We introduce a typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$ to
extract additional information from the static structure of terms in WHNF:
\[
    \TypeWHNFCons
\qquad
    \TypeWHNFLam
\]
This judgement differs from the main typing judgement in that (1) it only
applies to expressions in weak head normal form, and (2) it ``outputs'' (to the right of $\gtrdot$) a
disjoint set of linear environments ($\ov{\D_i}$), where each environment corresponds to the
linear resources used by a sub-expression of the WHNF expression.
%
To type a constructor application $K~\ov{e_\omega e_i}$, where $e_\omega$
are the unrestricted arguments and $e_i$ the linear arguments of the
constructor, we split the resources $\D$ into a disjoint set of resources
$\ov{\D_i}$ required to type each linear argument individually and return exactly
that split of the resources; the unrestricted $e_\omega$ expressions must be
typed on an empty linear environment. A lambda expression is typed with
the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
as there is always only a single sub-expression in lambda abstractions.

@@

%%%%???
%%%%The linear resources used to type an expression in WHNF are exactly those which
%%%%occur to the right of $\gtrdot$ in the WHNF judgement shown above
%%%%(corresponding to the resources required to typecheck the lambda body or the
%%%%constructor arguments).
